Everything I know about Subset Sum Reconfiguration

### 2013-07-16 notes/ssr.tex
Improve wording slightly

\begin{prop}
    Subset sum reconfiguration is weakly NP-hard. \textup(Demaine--Ito\textup)
\end{prop}
\begin{proof}
    The ordinary subset sum problem may be reduced to subset sum reconfiguration. Given elements
    $a_1, \dots, a_m$ with target sum $t$, build an SSR instance with elements $a_1, \dots, a_m$,
    $t_1$ and $t_2$, threshold values $k=t$ and $c=2t$, starting configuration $t_1$ and ending configuration
    $t_2$. At some point element $t_1$ must move from inside to outside: the first time it does, the
    inside set during the move must consist of some collection of $a$ elements that sum to $t$.
\end{proof}

@@

\begin{prop}\label{prop:element-moves-a-lot}
    For all $m$, there is a $(5m+3)$-element SSR instance that has an element that must move at least $2m+1$ times in any solution sequence.
\end{prop}
\begin{proof}
    The elements are $x_1, \dots, x_{2m+1}$ of weight $3$, $a_1, \dots, a_{3m+1}$ of weight $2$, and $b$ of weight $1$.
    Let $k=6m$ and $c=6m+3$. The starting configuration is $x_1, \dots, x_{2m+1}$, and the target is $a_1, \dots, a_{3m+1}, b$. For $i=0, 1, \dots, 2m$, consider the first move after which there are precisely $2m-i$ of the $x$ elements inside. That means the $a$'s and possibly $b$ inside at this point must sum to $3i$. If $i$ is odd then, since $b$ is the only element with odd weight, $b$ must be inside at this point; similarly if $i$ is even then $b$ must be outside. So element $b$ moves at least $2m$ times. Since it must end up inside, it must move an odd number of times altogether, so at least $2m+1$ times.
\end{proof}

\section*{New finding}
\noindent The construction of Proposition~\ref{prop:element-moves-a-lot} can be modified to give a quadratic lower bound for the diameter of the reconfiguration graph.
\begin{prop}
    For all $m$, there is a $(6m+2)$-element SSR instance that requires $2m^2+6m+2$ moves.
\end{prop}
\begin{proof}
    The elements are
    $x_1, \dots, x_{2m+1}$ of weight $3m$,
    $a_1, \dots, a_{3m+1}$ of weight $2m$,
    and $b_1, \dots, b_m$ of weight $1$.
    %
    Let $k=6m^2$ and $c=6m^2+3m$.
    %
    The starting configuration is $x_1, \dots, x_{2m+1}$,
    and the target is $a_1, \dots, a_{3m+1}, b_1, \dots, b_m$.
    %
    For $i=0, 1, \dots, 2m$, consider the first move after which there are precisely $2m-i$ of the $x$ elements inside. That means the $a$'s and $b$'s inside at this point must sum to $3im$. If $i$ is odd then, since the $a$'s all weigh $2m$, all the $b$'s must be inside at this point; similarly if $i$ is even then the $b$'s must all be outside. So the $b$ elements each move at least $2m$ times, and in fact $2m+1$ times since they move.
    %
    Therefore the total number of moves must be at least $(2m+1) + (3m+1) + (2m+1)m$, which is equal to $2m^2+6m+2$ as required.
\end{proof}
\end{document}

>>>>

\begin{prop}
    Subset sum reconfiguration is weakly NP-hard. \textup(Demaine--Ito\textup)
\end{prop}
\begin{proof}
    The ordinary subset sum problem may be reduced to subset sum reconfiguration. Given elements
    $a_1, \dots, a_m$ with target sum $t$, build an SSR instance with elements $a_1, \dots, a_m$ having the corresponding weights,
    $t_1$ and $t_2$ having weight $t$, threshold values $k=t$ and $c=2t$, starting configuration $t_1$ and ending configuration
    $t_2$. At some point element $t_1$ must move from inside to outside: the first time it does, the
    inside set during the move must consist of some collection of $a$ elements that sum to $t$.
\end{proof}

@@

\begin{prop}\label{prop:element-moves-a-lot}
    For all $m$, there is a $(5m+3)$-element SSR instance that has an element that must move at least $2m+1$ times in any solution sequence.
\end{prop}
\begin{proof}
    The elements are $x_1, \dots, x_{2m+1}$ of weight $3$, $a_1, \dots, a_{3m+1}$ of weight $2$, and $b$ of weight $1$.
    Let $k=6m$ and $c=6m+3$. The starting configuration is $x_1, \dots, x_{2m+1}$, and the target is $a_1, \dots, a_{3m+1}, b$. For $i=0, 1, \dots, 2m$, consider the first move after which there are precisely $2m-i$ of the $x$ elements inside. That means the $a$'s and possibly $b$ inside at this point must sum to $3i$. If $i$ is odd then, since the $a$'s all have even weight, $b$ must be inside at this point; similarly if $i$ is even then $b$ must be outside. So element $b$ moves at least $2m+1$ times.
\end{proof}

\section*{New finding}
\noindent The construction of Proposition~\ref{prop:element-moves-a-lot} can be modified to give a quadratic lower bound for the diameter of the reconfiguration graph.
\begin{prop}
    For all $m$, there is a $(6m+2)$-element SSR instance that requires $2m^2+6m+2$ moves.
\end{prop}
\begin{proof}
    The elements are
    $x_1, \dots, x_{2m+1}$ of weight $3m$,
    $a_1, \dots, a_{3m+1}$ of weight $2m$,
    and $b_1, \dots, b_m$ of weight $1$.
    %
    Let $k=6m^2$ and $c=6m^2+3m$.
    %
    The starting configuration is $x_1, \dots, x_{2m+1}$,
    and the target is $a_1, \dots, a_{3m+1}, b_1, \dots, b_m$.
    %
    For $i=0, 1, \dots, 2m$, consider the first move after which there are precisely $2m-i$ of the $x$ elements inside. That means the $a$'s and $b$'s inside at this point must sum to $3im$. If $i$ is odd then, since the $a$'s all weigh $2m$, all the $b$'s must be inside at this point; similarly if $i$ is even then the $b$'s must all be outside. So the $b$ elements each move at least $2m+1$ times.
    %
    Therefore the total number of moves must be at least $(2m+1) + (3m+1) + (2m+1)m$, which is equal to $2m^2+6m+2$ as required.
\end{proof}
\end{document}
